\begin{abstract}

This thesis explores the applicability of Kubernetes as a platform for implementing business logic using Custom Resource Definitions (CRDs) and the Operator pattern. While Kubernetes has become the de facto standard for container orchestration, its potential as a business logic platform remains understudied.

Two functionally equivalent billing systems were designed and implemented: (1) a traditional Go backend with REST API and PostgreSQL database, and (2) a Kubernetes Operator with CRDs and reconcile-based control loop. Both systems implement identical business logic (billing plan management, subscription lifecycle, recurring billing, expiration handling) but differ fundamentally in architecture, state management, and execution model.

Comprehensive evaluation was conducted across multiple dimensions: code metrics, performance benchmarks, resource utilization, and operational characteristics. The results demonstrate that the traditional backend achieves 10--20$\times$ lower latency (p50: 10ms vs 100ms for subscription creation) and 5$\times$ higher throughput (~1,000 req/s vs ~200 req/s). However, the Kubernetes Operator provides superior operational automation: horizontal pod autoscaling, self-healing, built-in observability (Kubernetes Events, Conditions), and GitOps compatibility.

The central claim is that Kubernetes Operators constitute a lifecycle-oriented business state management platform --- a paradigm distinct from both REST microservices and infrastructure-only operators --- and that the cost of choosing it for business logic is empirically quantifiable rather than anecdotal. The accompanying decision framework recommends the traditional approach for request-driven workloads where the natural unit of work completes synchronously, and the operator approach where the natural unit of work is the lifecycle of a long-lived entity with structured state transitions, guaranteed cleanup hooks, and an audit-trail requirement.

Both implementations are released as open source and are presented as reference rather than production code; outstanding limitations (no payment-gateway integration, currency-agnostic revenue metric, single-replica worker, asymmetric lifecycle vocabularies between the two systems) are catalogued in the conclusion. The findings suggest that Kubernetes can serve as a viable business-logic platform when business workflows are naturally expressible as resource lifecycles, but that the latency-automation trade-off should be evaluated explicitly against organisational context, team expertise, and the specific shape of the workload, rather than treated as a technological default.

\end{abstract}

\begin{abstract}[russian]

Данная диссертация исследует применимость Kubernetes в качестве платформы для реализации бизнес-логики с использованием Custom Resource Definitions (CRD) и паттерна Operator. Несмотря на то, что Kubernetes стал стандартом де-факто для оркестрации контейнеров, его потенциал как платформы для бизнес-логики остаётся недостаточно изученным.

Были спроектированы и реализованы две функционально эквивалентные биллинг-системы: (1) традиционный Go-бэкенд с REST API и базой данных PostgreSQL, (2) Kubernetes Operator с CRD и циклом reconcile. Обе системы реализуют идентичную бизнес-логику (управление тарифными планами, жизненный цикл подписок, периодический биллинг, обработка истечения срока), но фундаментально различаются архитектурой, управлением состоянием и моделью выполнения.

Проведено комплексное сравнение по нескольким измерениям: метрики кода, бенчмарки производительности, использование ресурсов и операционные характеристики. Результаты показывают, что традиционный бэкенд обеспечивает в 10--20 раз меньшую задержку (p50: 10 мс против 100 мс для создания подписки) и в 5 раз более высокую пропускную способность (~1000 запросов/с против ~200 запросов/с). Однако Kubernetes Operator предоставляет превосходную операционную автоматизацию: автоматическое масштабирование (HPA), самовосстановление, встроенную наблюдаемость (Kubernetes Events, Conditions) и совместимость с GitOps.

Центральное утверждение работы состоит в том, что Kubernetes Operators представляют собой платформу для управления состоянием жизненного цикла бизнес-сущностей --- парадигму, отличную как от REST-микросервисов, так и от операторов инфраструктурного назначения, --- а стоимость её применения для бизнес-логики поддаётся эмпирическому измерению, а не оценивается анекдотически. Сопровождающая рамка принятия решений рекомендует традиционный подход для нагрузок, естественной единицей работы которых является синхронный запрос, и операторный подход для нагрузок, естественной единицей работы которых является жизненный цикл долгоживущей сущности с явными переходами состояний, гарантированными точками очистки и требованием журнала аудита.

Обе реализации выпущены как open-source в качестве референсного, а не production-кода; остающиеся ограничения (отсутствие интеграции с платёжным шлюзом, отсутствие меток валюты в метрике выручки, отсутствие leader-election в фоновом воркере, асимметрия словарей состояний между двумя системами) систематизированы в заключении. Выводы показывают, что Kubernetes может служить жизнеспособной платформой для бизнес-логики, когда бизнес-процессы естественно выражаются через жизненный цикл ресурсов, но компромисс между задержкой и автоматизацией следует оценивать применительно к организационному контексту, экспертизе команды и конкретной форме нагрузки, а не принимать как технологическую данность.

\end{abstract}
