\chapter{Methodology}\label{chap:met}

This chapter describes the research methodology, system design, and architectural decisions used to evaluate Kubernetes as a platform for implementing business logic. The chapter expands upon the high-level methodology outlined in Chapter~\ref{chap:lr} and Section~1.5, providing a detailed description of both the Kubernetes-based system and the traditional baseline system.

\section{Research Approach}

This research followed a system design and comparative analysis approach. Two functionally equivalent systems were designed and implemented:

\begin{itemize}
    \item A Kubernetes-native system based on Custom Resource Definitions (CRDs) and the Operator pattern.
    \item A traditional backend system based on a REST API and relational database.
\end{itemize}

Both systems implemented the same business domain logic to ensure comparability. The primary difference lies in how the business logic is modeled, executed, and managed.

The methodology focuses on analyzing:
\begin{itemize}
    \item Architectural differences between declarative and imperative approaches.
    \item Execution models (event-driven reconciliation vs request-response).
    \item Observability and system introspection capabilities.
\end{itemize}

\section{Business Domain Modeling}

The selected domain for this research is a billing and subscription management system. This domain was chosen because it includes both stateful entities and time-based business logic, making it suitable for evaluating different architectural paradigms.

\subsection{Core Entities}

The system defines two primary entities:

\begin{itemize}
    \item \textbf{BillingPlan}: Represents a pricing tier, including attributes such as price, currency, billing period, and usage limits.
    \item \textbf{Subscription}: Represents a user's subscription to a billing plan, including references to the plan, lifecycle state, and billing timestamps.
\end{itemize}

\subsection{Business Logic}

The implemented business logic includes:

\begin{itemize}
    \item \textbf{Activation}: Upon creation of a subscription, the system validates the referenced billing plan, activates the subscription, and initializes billing timestamps.
    \item \textbf{Billing Cycle Management}: The system tracks last payment and next billing dates.
    \item \textbf{Expiration Handling}: Subscriptions transition to an expired state when billing conditions are not met.
    \item \textbf{Final Charge Handling}: Upon deletion, a final billing operation is executed before resource removal.
\end{itemize}

This logic was implemented identically in both systems but using different architectural approaches.

\section{Traditional System Architecture}

The baseline system follows a conventional monolithic architecture.

\subsection{Technology Stack}

\begin{itemize}
    \item Programming language: Go
    \item Web framework: Gorilla Mux
    \item Database: PostgreSQL 15
    \item ORM: GORM
\end{itemize}

\subsection{Architecture Overview}

The system exposes a REST API that allows clients to interact with billing plans and subscriptions. Business logic is implemented within HTTP handlers and database operations.

\subsection{Execution Model}

The system uses a primarily synchronous request-response model:

\begin{itemize}
    \item Clients send HTTP requests to perform operations.
    \item The server processes requests and updates the database.
\end{itemize}

Additionally, an asynchronous background worker periodically checks for expired subscriptions and updates their state accordingly.

\subsection{Data Persistence}

All system state is stored in a relational database. The database acts as the single source of truth, and business logic is tightly coupled with database operations.

\section{Kubernetes-Based System Architecture}

The experimental system is implemented using Kubernetes as a platform for business logic.

\subsection{Custom Resource Definitions}

Two CRDs were defined:

\begin{itemize}
    \item \texttt{BillingPlan} (\texttt{billingplans.billing.cloud-native.io})
    \item \texttt{Subscription} (\texttt{subscriptions.billing.cloud-native.io})
\end{itemize}

These CRDs extend the Kubernetes API, allowing domain entities to be managed as native Kubernetes resources.

\subsection{Operator Implementation}

The system follows the Operator pattern and is implemented using Kubebuilder and controller-runtime.

The core logic resides in the \texttt{SubscriptionReconciler}, which continuously reconciles the desired and actual state of Subscription resources.

\subsection{Reconciliation Logic}

The reconciliation loop implements the following logic:

\begin{itemize}
    \item Validation of referenced BillingPlan existence.
    \item Automatic attachment of finalizers for safe deletion.
    \item Initialization of subscription state upon creation.
    \item Handling of deletion events with final billing logic.
    \item Continuous state synchronization.
\end{itemize}

The reconciliation loop effectively acts as a state machine, continuously driving the system toward the desired state.

\subsection{Execution Model}

Unlike the traditional system, execution is event-driven:

\begin{itemize}
    \item Changes to resources trigger reconciliation events.
    \item Controllers asynchronously process these events.
    \item No direct client-driven execution of business logic is required.
\end{itemize}

\subsection{Data Persistence}

System state is stored in the Kubernetes control plane (etcd) as custom resources. The state is declarative and versioned, enabling automatic reconciliation and recovery.

\section{Comparison of Architectural Approaches}

The two systems differ fundamentally in how they implement business logic:

\begin{itemize}
    \item \textbf{Imperative vs Declarative}: The REST system executes explicit commands, while the Kubernetes system declares desired state.
    \item \textbf{Execution Model}: Request-response vs reconciliation loop.
    \item \textbf{State Management}: Database records vs Kubernetes resources.
    \item \textbf{Error Handling}: Explicit error handling vs eventual consistency and retries.
\end{itemize}

These differences form the basis for the evaluation conducted in later chapters.

\section{Observability Design}

Observability was implemented in both systems to enable analysis of system behavior.

\subsection{Kubernetes System}

The Kubernetes-based system integrates both metrics and tracing:

\begin{itemize}
    \item \textbf{Metrics (Prometheus)}:
    \begin{itemize}
        \item \texttt{billing\_active\_subscriptions}
        \item \texttt{billing\_revenue\_total}
        \item \texttt{billing\_payment\_failures}
    \end{itemize}
    \item \textbf{Tracing (OpenTelemetry)}:
    \begin{itemize}
        \item Tracing of reconciliation events via spans (e.g., \texttt{ReconcileSubscription})
    \end{itemize}
\end{itemize}

This setup enables visibility into business-level events rather than only infrastructure metrics.

\subsection{Traditional System}

The traditional system uses:

\begin{itemize}
    \item Standard HTTP metrics (request latency, request count)
    \item OpenTelemetry tracing integrated with database operations
\end{itemize}

Tracing captures the flow from API requests to database interactions.

\section{Experimental Methodology}

Experimental evaluation will be conducted in a controlled local environment using a Kubernetes cluster (Minikube).

The evaluation will focus on:

\begin{itemize}
    \item Reaction time to state changes
    \item Throughput under load
    \item Resource utilization
    \item Observability coverage
\end{itemize}

Detailed experimental setup and results are presented in Chapter~\ref{chap:eval}.

\section{Summary}

This chapter described the methodology used to evaluate Kubernetes as a platform for business logic. Two systems implementing identical functionality were designed using different paradigms: a traditional REST-based system and a Kubernetes-native Operator-based system. The architectural differences between these systems provide the foundation for the comparative evaluation presented in subsequent chapters.
