\chapter{Conclusion}\label{chap:conclusion}

This chapter summarizes the research findings, answers the research questions posed in Chapter~\ref{chap:intro}, discusses the contributions and limitations of this work, and suggests directions for future research.

\section{Summary of Research}

This thesis explored the applicability of Kubernetes as a platform for implementing business logic using Custom Resource Definitions (CRDs) and the Operator pattern. Two functionally equivalent billing systems were designed, implemented, and evaluated:

\begin{itemize}
    \item \textbf{backend-billing}: A traditional Go backend with REST API and PostgreSQL database
    \item \textbf{kube-billing}: A Kubernetes Operator with CRDs and reconcile-based control loop
\end{itemize}

Both systems implement identical business logic (billing plan management, subscription lifecycle, recurring billing, expiration handling) but differ fundamentally in architecture, state management, and execution model.

\section{Answers to Research Questions}

\subsection{RQ1: How can business logic be effectively modeled and implemented using Kubernetes CRDs and the Operator pattern?}

\textbf{Answer}: Business logic can be modeled using Kubernetes CRDs by:

\begin{enumerate}
    \item \textbf{Defining domain entities as CRDs}: BillingPlan and Subscription resources extend the Kubernetes API with domain-specific semantics
    \item \textbf{Using spec/status separation}: Desired state (spec) is declared by users; actual state (status) is managed by the controller
    \item \textbf{Implementing reconcile loop}: The Operator continuously compares spec vs status and applies changes to converge toward desired state
    \item \textbf{Leveraging Kubernetes primitives}: Finalizers for cleanup, Conditions for state reporting, Events for audit trail
    \item \textbf{Using admission control}: CRD validation markers enforce business rules at API boundary
\end{enumerate}

The key insight is that business logic becomes a \textit{control problem} rather than a \textit{computation problem}: the Operator acts as an automated control system that maintains system state according to business rules.

\subsection{RQ2: What are the performance characteristics of CRD/Operator-based implementation compared to traditional REST API architectures?}

\textbf{Answer}: The experimental evaluation (Chapter~\ref{chap:eval}) demonstrates significant performance differences:

\begin{table}[h]
\centering
\caption{Performance Comparison Summary}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Metric} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Latency (p50, create subscription) & 10 ms & 100 ms \\
Throughput (requests/second) & $\sim$1,000 & $\sim$200 \\
CPU usage (idle) & 10m & 50m \\
Memory usage (idle) & 50 MB & 150 MB \\
Startup time & 2s & 10s \\ \bottomrule
\end{tabularx}
\end{table}

The traditional backend is \textbf{10--20$\times$ faster} in latency and \textbf{5$\times$ higher} in throughput. The Kubernetes API server and etcd introduce overhead from admission control, distributed consensus, and watch event propagation.

\textbf{Conclusion}: For performance-critical applications, the traditional approach is superior. The Operator pattern trades raw performance for operational benefits.

\subsection{RQ3: What are the operational implications and observability capabilities of using Kubernetes as a business logic platform?}

\textbf{Answer}: The Kubernetes approach provides significant operational advantages:

\begin{itemize}
    \item \textbf{Automatic scaling}: HorizontalPodAutoscaler adjusts replicas based on CPU/memory utilization
    \item \textbf{Self-healing}: Failed pods are automatically restarted by Kubernetes
    \item \textbf{Declarative deployment}: Kustomize/Helm enable version-controlled, reproducible deployments
    \item \textbf{Built-in observability}:
    \begin{itemize}
        \item Kubernetes Events provide audit trail without custom logging
        \item Conditions (metav1.Condition) standardize state reporting
        \item Reconciliation metrics expose control loop behavior
    \end{itemize}
    \item \textbf{GitOps compatibility}: CRD manifests integrate with ArgoCD/Flux workflows
\end{itemize}

However, these benefits come with operational complexity:
\begin{itemize}
    \item Kubernetes cluster management required
    \item etcd backup and disaster recovery planning
    \item RBAC configuration for least-privilege access
    \item Monitoring of control plane health
\end{itemize}

\subsection{RQ4: Under what circumstances is the CRD/Operator approach advantageous, and what are its limitations?}

\textbf{Answer}: The CRD/Operator approach is advantageous when:

\begin{itemize}
    \item \textbf{Application is already on Kubernetes} --- no additional infrastructure investment
    \item \textbf{High availability is critical} --- self-healing and HPA provide resilience
    \item \textbf{Business logic is state machine-like} --- reconcile pattern fits naturally
    \item \textbf{Audit trail is required} --- Kubernetes Events provide compliance logging
    \item \textbf{GitOps workflow is desired} --- declarative resources enable PR-based changes
\end{itemize}

The approach is \textbf{not recommended} when:

\begin{itemize}
    \item \textbf{Performance is the top priority} --- 10--20$\times$ latency overhead
    \item \textbf{Team lacks Kubernetes expertise} --- steep learning curve
    \item \textbf{Application must run outside Kubernetes} --- portability limitation
    \item \textbf{Complex queries are needed} --- Kubernetes API is not designed for ad-hoc queries
    \item \textbf{Time-to-market is critical} --- slower development cycle
\end{itemize}

\section{Contributions}

This thesis makes the following contributions:

\begin{enumerate}
    \item \textbf{Two reference implementations released as open source}:
    \begin{itemize}
        \item backend-billing: Go REST API with PostgreSQL, OpenTelemetry instrumentation, Docker-Compose deployment with Jaeger and Prometheus, and a test suite covering apierrors, middleware, and validator packages (per-package coverage reported in Section~\ref{sec:coverage-results}).
        \item kube-billing: Kubernetes Operator with two CRDs, finalizer-driven pro-rated refund, controller-runtime metrics integration, OpenTelemetry tracing, HPA, PDB, and a Kustomize-based deployment overlay.
    \end{itemize}
    Both are reference quality rather than production quality: they have no payment-gateway integration, no horizontal-replica leader election in the traditional worker, and the operator's revenue counter is currency-agnostic (a single dimensionless metric). Section~\ref{sec:limitations} catalogues these and other limitations systematically.
    
    \item \textbf{Systematic architectural comparison}:
    \begin{itemize}
        \item Code metrics (LOC, complexity, dependencies, test coverage)
        \item Performance benchmarks (latency, throughput, resource usage)
        \item Operational characteristics (scaling, fault tolerance, observability)
    \end{itemize}
    
    \item \textbf{Decision framework}: Criteria for choosing between traditional and Operator approaches based on organizational priorities
    
    \item \textbf{Reference implementation}: Open-source codebases serving as templates for similar projects
\end{enumerate}

\section{Limitations}\label{sec:limitations}

The limitations of this work fall into four categories, mirroring the validity dimensions used in the evaluation chapter.

\paragraph{Implementation limitations.}
\begin{itemize}
    \item \textbf{No payment-gateway integration.} Both systems simulate payments by incrementing counters; real-world latency, idempotency keys, and partial-failure semantics of services such as Stripe are not captured.
    \item \textbf{Currency-agnostic revenue metric.} The kube-billing \texttt{billing\_revenue\_total} counter accumulates numerical price values without a currency label, so revenue across different currencies is conflated.
    \item \textbf{Hardcoded backend billing period.} The traditional handler computes the next billing date as \texttt{now + one month} regardless of the plan's declared \texttt{billingPeriod}, which means the systems do not interpret all plan configurations identically.
    \item \textbf{Single-replica worker.} The traditional expiration checker has no leader-election support, so the system cannot be safely horizontally scaled without external coordination.
    \item \textbf{Unused custom validators.} The traditional backend's \texttt{currency}, \texttt{billing\_period}, and \texttt{price} validators are registered but never invoked, leaving validation looser than the CRD's admission-time schema.
\end{itemize}

\paragraph{Methodological limitations.}
\begin{itemize}
    \item \textbf{Local single-node environment.} Experiments were conducted on a single Apple Silicon laptop running Minikube and Docker Desktop. Production clusters incur additional network latency to the API server, etcd raft latency, and apiserver flow-control behaviour that this environment cannot reproduce.
    \item \textbf{No failure injection.} The fault-tolerance claims were not exercised against forced pod kills, network partitions, etcd unavailability, or PostgreSQL primary failover.
    \item \textbf{Short measurement horizon.} Benchmarks ran for minutes, not hours or days; long-running effects (etcd compaction backlog, PostgreSQL index bloat, Prometheus cardinality growth) are not represented.
\end{itemize}

\paragraph{Internal validity threats.}
\begin{itemize}
    \item \textbf{Construct mismatch between throughput definitions.} backend-billing throughput is measured with concurrent HTTP load; kube-billing throughput is bounded by the time it takes the apiserver and reconciler to drive a resource to \texttt{status.state == Active}. The two measurements describe different system contracts and should not be subtracted as if they were comparable units.
    \item \textbf{Asymmetric workloads.} The two implementations do not expose identical state vocabularies (Table~\ref{tab:state-comparison}); the comparison is between systems that satisfy similar but not identical functional contracts.
\end{itemize}

\paragraph{External validity threats.}
\begin{itemize}
    \item \textbf{Single business domain.} Billing exhibits stateful, time-driven semantics; the conclusions may not transfer to request-driven workloads (search, content delivery) or pure data-pipeline workloads (ETL, streaming aggregation).
    \item \textbf{Single hardware platform.} Apple M3 unified memory and ARM64 microarchitecture differ from x86 server-class processors; absolute numbers may not generalise.
    \item \textbf{No multi-tenant evaluation.} Namespace-based isolation, resource quotas, and noisy-neighbour effects were not assessed.
\end{itemize}

\section{Future Work}

Several directions for future research emerge from this work:

\subsection{Technical Improvements}

\begin{itemize}
    \item \textbf{Payment gateway integration}: Implement Stripe/PayPal integration for realistic payment latency and failure handling
    
    \item \textbf{Refund logic}: Complete finalizer implementation with pro-rated refund calculation
    
    \item \textbf{Multi-tenant support}: Evaluate namespace-based isolation and resource quotas
    
    \item \textbf{Webhook integration}: Add ValidationWebhook and MutationWebhook for advanced admission control
    
    \item \textbf{Backup/restore}: Implement etcd backup strategies and disaster recovery procedures
\end{itemize}

\subsection{Research Extensions}

\begin{itemize}
    \item \textbf{Alternative domains}: Apply Operator pattern to other business domains (inventory management, order processing, user provisioning)
    
    \item \textbf{Hybrid architectures}: Explore combinations (e.g., REST API for user-facing operations, Operator for background processing)
    
    \item \textbf{Cost analysis}: Compare cloud infrastructure costs (managed Kubernetes vs managed database + compute)
    
    \item \textbf{Developer productivity}: Measure development velocity, onboarding time, and maintenance effort
    
    \item \textbf{Security analysis}: Evaluate RBAC configurations, secret management, and network policy effectiveness
\end{itemize}

\subsection{Production Deployment}

\begin{itemize}
    \item \textbf{Helm chart}: Package kube-billing as a Helm chart for easier deployment
    
    \item \textbf{Operator Lifecycle Manager}: Publish to OperatorHub for community distribution
    
    \item \textbf{Performance tuning}: Optimize etcd configuration, reconcile concurrency, and cache settings
    
    \item \textbf{Monitoring dashboards}: Create Grafana dashboards for business metrics visualization
\end{itemize}

\section{Final Remarks}

The central claim of this thesis is that Kubernetes Operators constitute a lifecycle-oriented business state management platform: a paradigm distinct from REST microservices, which centre on request-response semantics, and from infrastructure operators, which centre on resource provisioning. The case for this paradigm rests not on architectural fashion but on measurable trade-offs --- the reconcile loop and its surrounding primitives buy operational properties (eventual consistency, self-healing, finalizer-guarded cleanup, structured Conditions, declarative audit trail) at the price of an order-of-magnitude latency penalty and a multiple of the resource footprint compared with a conventional REST stack.

For practitioners, this implies a decision framework anchored in workload shape rather than in technological alignment:

\begin{itemize}
    \item If the natural unit of work is a request that must complete synchronously and the business does not need lifecycle hooks beyond cancellation, a traditional backend remains the right choice; the operator pattern adds latency, dependencies, and operational surface area without compensating benefit.
    \item If the natural unit of work is the lifecycle of a long-lived entity --- a subscription, a tenant, a content workflow --- and the business needs guaranteed cleanup, structured state reporting, and an audit trail that survives the absence of a human operator, the operator pattern's primitives genuinely earn their cost.
\end{itemize}

The contribution is methodological as well as empirical: by treating asymmetries (cancellation-as-deletion vs cancellation-as-state, error-as-Condition vs error-as-HTTP-status) as findings rather than as inconveniences to be smoothed away, the comparison surfaces what the two paradigms actually disagree about, which is what a practitioner needs to evaluate a choice that the surface-level feature matrices cannot.

\section{Publications and Artifacts}

The following artifacts are available as open-source:

\begin{itemize}
    \item \textbf{backend-billing}: \url{https://github.com/cuprum-acid/backend-billing}
    \item \textbf{kube-billing}: \url{https://github.com/cuprum-acid/kube-billing}
    \item \textbf{Thesis source}: \url{https://github.com/cuprum-acid/thesis}
\end{itemize}
