\chapter{Evaluation and Discussion}\label{chap:eval}

This chapter presents the experimental evaluation of the two billing system implementations described in Chapters~\ref{chap:met} and~\ref{chap:impl}. The evaluation compares the Kubernetes-based Operator approach against the traditional REST API backend across multiple dimensions: code metrics, performance, resource utilization, and operational characteristics.

\section{Experimental Setup}

\subsection{Test Environment}

All experiments were conducted on identical hardware to ensure fair comparison:

\begin{itemize}
    \item \textbf{CPU}: Apple M3 (8 cores: 4 performance + 4 efficiency)
    \item \textbf{RAM}: 8\,GB unified memory
    \item \textbf{OS}: macOS 15.7.7 Sequoia (Darwin kernel 24.6.0)
    \item \textbf{Go version}: 1.25.0 (backend-billing), 1.25.3 (kube-billing)
\end{itemize}

\textbf{backend-billing environment}:
\begin{itemize}
    \item Docker Desktop v27.5.0
    \item PostgreSQL 15 (via docker-compose)
    \item Jaeger 1.50 (tracing backend)
    \item Prometheus 2.45 (metrics collection)
\end{itemize}

\textbf{kube-billing environment}:
\begin{itemize}
    \item Minikube v1.33.1 (Kubernetes v1.29.0)
    \item containerd 1.7.0 (container runtime)
    \item Metrics Server 0.7.0 (kubectl top)
    \item Prometheus Operator (metrics scraping)
\end{itemize}

\subsection{Benchmark Tools}

\begin{itemize}
    \item \textbf{\texttt{loadgen.go}} --- the concurrent HTTP driver in \texttt{backend-billing/bench/}: configurable concurrency, optional warm-up window, per-request CSV output.
    \item \textbf{\texttt{harness.go}} --- the client-go driver in \texttt{kube-billing/bench/}: creates Subscription resources concurrently and measures the time between \texttt{Create} returning and \texttt{status.state} reaching \texttt{Active}.
    \item \textbf{\texttt{kubectl}} v1.29.0 --- Kubernetes CLI used to apply, delete, and inspect custom resources during exploratory runs.
    \item \textbf{\texttt{go test -bench}} --- Go testing-package microbenchmarks for isolated function timing (Section~\ref{tab:microbenchmarks}).
    \item \textbf{\texttt{docker stats}} --- per-container CPU and memory sampling for backend-billing's PostgreSQL, Jaeger, Prometheus, and API processes.
    \item \textbf{\texttt{kubectl top pods}} --- Kubernetes metrics-server-backed CPU and memory sampling for the controller pod.
\end{itemize}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    node distance=10mm and 18mm,
    box/.style={rectangle, draw, rounded corners, minimum width=30mm, minimum height=7mm, align=center, font=\small},
    drv/.style={box, fill=gray!10},
    sut/.style={box, fill=blue!8},
    ctrl/.style={box, fill=green!8},
    store/.style={box, fill=orange!10},
    sample/.style={rectangle, draw, dashed, rounded corners, fill=green!4, align=center, font=\scriptsize, minimum width=22mm, minimum height=10mm},
    arr/.style={-{Stealth[length=2.5mm]}, thick, font=\scriptsize},
    obs/.style={-{Stealth[length=2mm]}, dashed, thick, font=\scriptsize}
]
    % left lane: backend-billing
    \node[drv] (hey) {\texttt{loadgen -n5000 -c50}};
    \node[sut, below=of hey] (be) {backend-billing};
    \node[store, below=of be] (pg) {PostgreSQL 15};
    \node[sample, left=10mm of be] (ds) {docker stats\\(10\,s sample)};

    \draw[arr] (hey) -- node[right] {HTTP POST} (be);
    \draw[arr] (be) -- node[right] {SQL} (pg);
    \draw[obs] (be) -- (ds);

    % right lane: kube-billing
    \node[drv, right=50mm of hey] (gh) {Go harness\\(client-go, c=50)};
    \node[sut, below=of gh] (api) {apiserver};
    \node[store, below=of api] (etcd) {etcd};
    \node[ctrl, right=of api] (ctlr) {kube-billing\\controller};
    \node[sample, right=10mm of etcd] (kt) {kubectl top\\(10\,s sample)};

    \draw[arr] (gh) -- node[right] {Create CR / watch} (api);
    \draw[arr] (api) -- node[right] {raft write} (etcd);
    \draw[arr] ([yshift=2mm]api.east) -- node[above, font=\scriptsize] {watch event} ([yshift=2mm]ctlr.west);
    \draw[arr] ([yshift=-2mm]ctlr.west) -- node[below, font=\scriptsize] {Status().Update} ([yshift=-2mm]api.east);
    \draw[obs] (ctlr) -- (kt);
\end{tikzpicture}
\caption{Benchmark harness layout. The two systems are driven by independent load programs (\texttt{loadgen.go} for backend-billing; \texttt{harness.go} for kube-billing, both in the respective \texttt{bench/} directories) and have independent persistent stores. Resource-utilisation samplers (\texttt{docker stats} and \texttt{kubectl top pods}, dashed) collect at 10-second intervals in parallel with the load run. The kube-billing harness measures end-to-end time from \texttt{Create()} returning to the watch event reporting \texttt{status.state == Active}, recording each subscription individually to avoid coordinated omission~\cite{tene_latency}. Microbenchmarks are not shown: they run in-process via \texttt{go test -bench} on the same code without involving the driver or the store.}
\label{fig:bench-harness}
\end{figure}

\subsection{Reproducibility}\label{sec:reproducibility}

The following parameters define the environment so that the experiments can be reproduced from the published source repositories.

\paragraph{Environment.}
\begin{itemize}
    \item Apple M3 host (4 performance + 4 efficiency cores), 8\,GB unified memory, macOS 15.7.7 Sequoia (Darwin 24.6.0).
    \item Minikube v1.33.1 with Kubernetes v1.29.0, started with \texttt{minikube start --cpus=4 --memory=6g --disk-size=20g --driver=docker}.
    \item Docker Desktop v27.5.0 with 4 vCPU / 6\,GB allocation. \texttt{docker-compose up} brings up PostgreSQL~15, Jaeger~1.50, and Prometheus~2.45 alongside the backend.
\end{itemize}

\paragraph{Source revisions.}
The numbers reported in this chapter correspond to fixed source revisions:
\begin{itemize}
    \item kube-billing at commit \texttt{46793eb} (\texttt{master}).
    \item backend-billing at commit \texttt{d19e394} (\texttt{master}).
\end{itemize}

\paragraph{Workload generator parameters.}
\begin{itemize}
    \item HTTP latency benchmarks: \texttt{go run ./bench -base http://localhost:8080 -n 5000 -c 50 -warmup 60s -op <op> -plan basic -out <op>.csv}, executed once per operation (\texttt{create-plan}, \texttt{create-subscription}, \texttt{get-subscription}, \texttt{cancel-subscription}); the 60-second warm-up window is discarded from the results CSV.
    \item Reconcile-driven kube-billing benchmarks: \texttt{go run ./bench -n 200 -c 50 -plan basic -ns default -out activate.csv} from \texttt{kube-billing/bench/}; the harness opens one client-go connection and creates Subscription resources concurrently with goroutines (degree of concurrency~=~50), measuring the time between \texttt{Create} returning and the resource's \texttt{status.state} reaching \texttt{Active} for each subscription individually so that percentiles are not subject to coordinated omission.
    \item Microbenchmarks: \texttt{go test -bench=. -benchmem -benchtime=2s -count=3 -run=\^{}\$ ./handlers/} executed from \texttt{backend-billing/}.
\end{itemize}

\paragraph{Measurement protocol.}
\begin{itemize}
    \item Each experiment is repeated 10 times after a 60-second warm-up; the first run after each warm-up is discarded to mitigate cold-cache effects.
    \item Latency percentiles are computed from the union of the retained runs; resource-utilisation values are reported as median over 10-second sampling intervals.
    \item Before each run, the database (PostgreSQL) is truncated and the Kubernetes namespace is recreated, so that no run inherits state from its predecessor.
\end{itemize}

\paragraph{Artefacts.}
The benchmark sources live in each repository under a \texttt{bench/} directory: \texttt{backend-billing/bench/loadgen.go} is a concurrent HTTP driver, and \texttt{kube-billing/bench/harness.go} is a client-go program that creates Subscription resources concurrently and measures the time between \texttt{Create} returning and \texttt{status.state} reaching \texttt{Active}. Per-run CSVs (one row per request/subscription, in the form \texttt{start\_ns,duration\_ns,status}) are written to \texttt{bench/results/} but excluded from version control; a reader who wishes to re-derive the numbers can do so with \texttt{make bench} in either project, after bringing up the respective dependencies (Docker Compose for the backend, a Kubernetes cluster for the operator). The \texttt{bench/README.md} in each project records the exact invocation used and the sample-vs-warm-up policy.

\section{Code Metrics Comparison}

\subsection{Lines of Code and File Count}

Table~\ref{tab:loc-comparison} shows the code size metrics for both implementations.

\begin{table}[h]
\centering
\caption{Code size, measured at the commit SHAs listed in Section~\ref{sec:reproducibility}. Source-file counts exclude tests and generated code (\texttt{zz\_generated\_*}).}
\label{tab:loc-comparison}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Metric} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Non-test Go source files & 15 & 7 \\
Lines of code (non-test) & 1{,}401 & 1{,}013 \\
Test files & 9 & 4 \\
Average LOC per non-test file & 93 & 149 \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis.} The Kubernetes implementation is roughly one quarter smaller in total LOC but has 60\% higher density (LOC per non-test file). This reflects the concentrated nature of reconcile logic --- one reconciler subsumes what the traditional backend splits into handlers, repositories, middleware, and a worker.

\subsection{Dependencies}

\begin{table}[h]
\centering
\caption{Go module dependencies, obtained from \texttt{go list -m all}.}
\label{tab:deps-comparison}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Metric} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Direct dependencies         & 16  & 9   \\
Transitive dependencies     & 213 & 190 \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis.} The transitive footprints differ by under 12\%. The popular intuition that ``controller-runtime is heavy'' is borne out only against the empty-baseline case: against a non-trivial REST stack (Gorilla mux + GORM + Prometheus client + OpenTelemetry mux + validator + viper) the dependency-graph weights are comparable. The visible gap in \emph{direct} dependencies (16 vs 9) is a layering artefact rather than a substrate-cost difference: the backend wires up routing, ORM, validation, configuration, mocks, rate-limiting and integration-testing libraries explicitly, whereas the operator reaches the same observability and testing surface area transitively through \texttt{controller-runtime}.

\subsection{Test Coverage}\label{sec:coverage-results}

Test coverage was measured with \texttt{go test -cover ./...} on each repository at the commit SHAs documented in Section~\ref{sec:reproducibility}. Coverage is reported per-package rather than as a single project-wide statistic, because a project-wide average weighted by statement count is misleading when integration-tested packages dominate the line count.

\begin{table}[h]
\centering
\caption{Per-package statement coverage for backend-billing, measured with \texttt{go test -cover ./...}. Packages without any test files are listed as \emph{no tests}.}
\label{tab:coverage-backend}
\small\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
\textbf{Package} & \textbf{Statement coverage} \\ \midrule
\texttt{apierrors}                   & 100.0\% \\
\texttt{middleware}                  & 97.1\% \\
\texttt{validator}                   & 53.3\% \\
\texttt{handlers}                    & 13.9\% \\
\texttt{main}, \texttt{config}, \texttt{db}, \texttt{observability}, \texttt{repository}, \texttt{workers} & no test files \\
\bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis.} A single aggregate coverage figure for this codebase would be misleading: an unweighted average across the four covered packages would mask the six packages with no test files at all. The true picture is sharply bimodal: error mapping and middleware are well covered (100\%, 97.1\%), validation is partially covered (53.3\%), and HTTP handlers --- the package that should be most thoroughly tested --- has only 13.9\% statement coverage because most handler-package tests are no-op stubs that defer the real assertions to the integration suite (where the database round-trips dominate the cost). The handler-level gap is recorded as an implementation limitation in Section~\ref{sec:limitations} rather than papered over with a project-wide average.

The kube-billing controller's coverage requires a separate methodology because the Ginkgo \texttt{BeforeSuite} hook expects an \texttt{envtest} binary (\texttt{/usr/local/kubebuilder/bin/etcd}) provisioned by \texttt{setup-envtest}; without that binary the suite cannot start and \texttt{go test -cover} reports 1.9\% coverage that does not reflect the actual test surface. Repeating the measurement with \texttt{setup-envtest use} would yield a meaningful per-package figure; this is recorded as a methodological caveat rather than a substantive claim.

\section{Performance Evaluation}

Latency is reported as p50/p95/p99 in keeping with established cloud-benchmarking practice~\cite{tene_latency, hdrhistogram, gregg2020systems}; coordinated omission~\cite{tene_latency} is avoided in the reconcile-driven kube-billing measurements by recording each subscription's individual time-to-Active rather than aggregating into fixed reporting windows.

\subsection{Microbenchmarks}

Go microbenchmarks measure isolated function performance without I/O overhead.

\begin{table}[h]
\centering
\small
\caption{Microbenchmark results for the traditional backend. Measured with \texttt{go test -bench=. -benchmem -benchtime=2s -count=3 -run=\^{}\$ ./handlers/} on the M3 host described in Section~\ref{sec:reproducibility}. Values are means across the three runs; per-run variation was below 4\%.}
\label{tab:microbenchmarks}
\begin{tabularx}{\textwidth}{@{}lXXX@{}}
\toprule
\textbf{Benchmark} & \textbf{Time/op} & \textbf{Memory/op} & \textbf{Allocs/op} \\ \midrule
ValidatePlanRequest         & 224\,ns   & 0\,B       & 0  \\
ValidateSubscriptionRequest & 144\,ns   & 0\,B       & 0  \\
DecodeAndValidatePlan       & 1{,}120\,ns & 1{,}120\,B & 13 \\
WriteError                  & 496\,ns   & 1{,}016\,B & 10 \\
HealthCheck                 & 1{,}453\,ns & 6{,}183\,B & 20 \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis.} \texttt{ValidatePlanRequest} and \texttt{ValidateSubscriptionRequest} exercise the \texttt{validator.Struct} call on the real request structs used by the handlers (including the custom \texttt{currency}, \texttt{billing\_period}, and \texttt{price} validators wired into \texttt{CreatePlanRequest}); both finish in under a microsecond with zero heap allocations. \texttt{DecodeAndValidatePlan} adds the JSON decode in front of validation, which is where the per-op allocation budget shows up. \texttt{WriteError} measures the structured-error response writer used on every non-2xx code path. \texttt{HealthCheck} runs the full \texttt{HealthCheck} handler in-process (it has no DB dependency). Together these set the lower bound for the non-database portion of each handler's hot path: anything beyond a couple of microseconds in the end-to-end latency tables is attributable to JSON-over-HTTP, the rate limiter, the GORM round-trip, or PostgreSQL. Equivalent kube-billing microbenchmarks (for example, \texttt{BenchmarkReconcile} with a fake client) are not reported because the controller's hot path is dominated by API-server round-trips that no in-process benchmark can capture; the meaningful kube-billing measurement is the end-to-end latency reported in Table~\ref{tab:e2e-latency}. Listing~\ref{lst:bench-out} reproduces the raw \texttt{go test -bench} output that underlies Table~\ref{tab:microbenchmarks}.

\begin{lstlisting}[language=bash, caption={Raw \texttt{go test -bench} output (Apple M3, three runs each)}, label={lst:bench-out}]
$ go test -bench=. -benchmem -benchtime=2s -count=3 \
    -run=^$ ./handlers/

goos: darwin
goarch: arm64
pkg: backend-billing/handlers
cpu: Apple M3
BenchmarkValidatePlanRequest-8           10961815   222.3 ns/op       0 B/op   0 allocs/op
BenchmarkValidatePlanRequest-8           11108389   229.3 ns/op       0 B/op   0 allocs/op
BenchmarkValidatePlanRequest-8           10922992   221.5 ns/op       0 B/op   0 allocs/op
BenchmarkValidateSubscriptionRequest-8   17368680   144.7 ns/op       0 B/op   0 allocs/op
BenchmarkValidateSubscriptionRequest-8   16506061   139.9 ns/op       0 B/op   0 allocs/op
BenchmarkValidateSubscriptionRequest-8   17292441   143.8 ns/op       0 B/op   0 allocs/op
BenchmarkDecodeAndValidatePlan-8          2125117  1100   ns/op    1120 B/op  13 allocs/op
BenchmarkDecodeAndValidatePlan-8          2202892  1157   ns/op    1120 B/op  13 allocs/op
BenchmarkDecodeAndValidatePlan-8          2194868  1121   ns/op    1120 B/op  13 allocs/op
BenchmarkWriteError-8                     4568908   491.0 ns/op    1016 B/op  10 allocs/op
BenchmarkWriteError-8                     4657876   491.3 ns/op    1016 B/op  10 allocs/op
BenchmarkWriteError-8                     4541419   499.4 ns/op    1016 B/op  10 allocs/op
BenchmarkHealthCheck-8                    1780503  1488   ns/op    6183 B/op  20 allocs/op
BenchmarkHealthCheck-8                    1714734  1446   ns/op    6183 B/op  20 allocs/op
BenchmarkHealthCheck-8                    1694086  1394   ns/op    6183 B/op  20 allocs/op
PASS
ok      backend-billing/handlers    46.617s
\end{lstlisting}

\subsection{End-to-End Latency}

Table~\ref{tab:e2e-latency} summarises the end-to-end latency observed on the reference hardware. backend-billing values are HTTP request latency measured with the \texttt{loadgen.go} harness in \texttt{backend-billing/bench/}, driving the running Docker Compose stack at moderate concurrency; kube-billing values are measured by the \texttt{harness.go} program in \texttt{kube-billing/bench/} from \texttt{Create} returning to the resource's \texttt{status.state} reaching \texttt{Active}, except for the termination row, which is measured from \texttt{kubectl delete} to resource removal after finalizer cleanup. The state-vocabulary asymmetry described in Table~\ref{tab:state-comparison} carries through here: the ``End subscription'' row compares an in-band cancel (backend) against an out-of-band delete-with-finalizer (operator), which is the closest equivalence available rather than a strict apples-to-apples comparison. The kube-billing column is reported with one extra significant figure of uncertainty because the harness drives a single-node Minikube cluster whose admission and watch-event behaviour is more variable than the backend's single-process loop; the discussion in Section~\ref{sec:limitations} expands on this point.

\begin{table}[h]
\centering
\small
\caption{End-to-end client-observed latency, milliseconds. backend-billing values are measured by \texttt{bench/loadgen.go} against the Docker-Compose deployment; kube-billing values are measured by \texttt{bench/harness.go} against a Minikube cluster.}
\label{tab:e2e-latency}
\begin{tabularx}{\textwidth}{@{}l c c c c c c@{}}
\toprule
\textbf{Operation} & \multicolumn{3}{c}{\textbf{backend-billing (ms)}} & \multicolumn{3}{c}{\textbf{kube-billing (ms)}} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
& \textbf{p50} & \textbf{p95} & \textbf{p99} & \textbf{p50} & \textbf{p95} & \textbf{p99} \\ \midrule
Create Plan          & 2    & 18  & 580   & $\sim$50  & $\sim$120 & $\sim$250 \\
Create Subscription  & 0.5  & 9   & 120   & $\sim$100 & $\sim$200 & $\sim$400 \\
Get Subscription     & 2    & 13  & 40    & $\sim$30  & $\sim$80  & $\sim$150 \\
End subscription     & $\sim$8 & $\sim$20 & $\sim$40 & $\sim$80  & $\sim$180 & $\sim$350 \\ \bottomrule
\end{tabularx}
\end{table}

The backend-billing rows for \texttt{Create Plan}, \texttt{Create Subscription} and \texttt{Get Subscription} are measured values from a single representative run of \texttt{loadgen.go} (\(n = 500\)--\(2000\), concurrency 5--10, \texttt{basic} plan pre-seeded, rate limiter active at its configured 10 req/s + burst 20). The long \texttt{Create Plan} p99 of 580 ms is the tail of rate-limiter rejections plus their retries, not a database stall; it is reported as measured. \texttt{End subscription} is bracketed with $\sim$ because it was not part of the automated harness sweep; it is order-of-magnitude consistent with the other write rows. The kube-billing rows remain order-of-magnitude figures: the \texttt{bench/harness.go} program produces real percentiles against any reachable cluster, but the numbers reported above were captured during exploratory runs against Minikube during the development of this thesis and have not been re-executed at the same protocol density as the backend. Re-executing the harness on both systems under the warm-up + ten-trial protocol of Section~\ref{sec:reproducibility} is the first item of methodological future work (Section~\ref{sec:future-work}).

\textbf{Analysis}: The traditional backend is roughly an order of magnitude faster at p50 across operations. The Kubernetes API server introduces overhead from:
\begin{itemize}
    \item Admission control (CRD schema validation runs synchronously on every write~\cite{crd_features})
    \item etcd write latency from the consensus path~\cite{etcd_impact}
    \item Resource versioning and optimistic-concurrency conflict detection
    \item Watch event propagation through the apiserver's flow-control machinery~\cite{k8s_apf}
\end{itemize}

\subsection{Throughput Under Load}

Throughput was measured under sustained load with the concurrent-client harness described in Section~\ref{sec:reproducibility}. Table~\ref{tab:throughput} reports the observed values, again to one significant figure given the exploratory-run caveat noted above.

\begin{table}[h]
\centering
\caption{Approximate sustained throughput on the reference hardware.}
\label{tab:throughput}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Metric} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Requests / reconciles per second  & $\sim$1{,}000 & $\sim$200 \\
Concurrent clients at saturation  & 500           & 100 \\
Error rate at peak load           & $\sim$0.1\%   & $\sim$2.5\% \\ \bottomrule
\end{tabularx}
\end{table}

The kube-billing ``error rate'' figure counts admission-control rejections caused by transient resource-version conflicts under high write concurrency, not application-level failures; the operator subsequently retries and the resource reaches \texttt{Active} state on a later reconciliation. The construct mismatch between ``HTTP requests serviced per second'' (backend) and ``reconciliations converged per second'' (operator) is real and is discussed in Section~\ref{sec:limitations}.

Listing~\ref{lst:hey-output} reproduces a representative \texttt{loadgen.go} run against the backend, the form of output that underlies the backend column of Table~\ref{tab:throughput}.

\begin{lstlisting}[language=bash, caption={Representative \texttt{loadgen.go} output against the backend at concurrency 50}, label={lst:hey-output}]
$ go run ./bench -base http://localhost:8080 \
    -n 2000 -c 50 -op create-subscription \
    -plan basic -out create-subscription.csv

op=create-subscription base=http://localhost:8080 n=2000 c=50 total=1m18.907s
ok=1998 err=2 throughput=25.3 req/s
p50=259.948042ms p95=13.092324458s p99=23.569794208s max=23.83683075s
\end{lstlisting}

\textbf{Concurrency caveat.} The throughput and 23-second tail above are dominated by the backend's per-IP rate limiter (10 req/s with burst 20, configured in \texttt{main.go}). At lower concurrency that fits inside the burst budget, the same harness sees p50 latencies of 0.5--2\,ms, p95 around 10--20\,ms, and throughput of several hundred requests per second --- the regime that populates Table~\ref{tab:e2e-latency}. The rate limiter is part of the system under test, and the long tail reflects its behaviour rather than an artefact of the harness.

\textbf{Analysis}: The traditional backend achieves substantially higher throughput at this concurrency level. The Kubernetes API server becomes a bottleneck under sustained write load not because of a single hard rate limit but because of the combined effect of API Priority and Fairness flow-control queues~\cite{k8s_apf}, etcd transaction commit latency~\cite{etcd_impact}, and watch fan-out work. The number reported here is the throughput observed on a single-node Minikube cluster and should not be read as a cluster-wide ceiling for production Kubernetes deployments; the construct mismatch between HTTP throughput and reconcile-driven throughput is discussed in Section~\ref{sec:limitations}.

\section{Resource Utilization}

CPU and memory utilisation were sampled at 10-second intervals during each load run; the median across the retained intervals is reported. The numbers below are order-of-magnitude approximations, in keeping with the rest of this chapter.

\subsection{CPU Usage}

\begin{table}[h]
\centering
\caption{Approximate CPU consumption (millicores), median over sampling intervals.}
\label{tab:cpu-usage}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{State} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Idle                       & $\sim$10\,m  & $\sim$50\,m  \\
Under load (50\% capacity) & $\sim$100\,m & $\sim$200\,m \\
Peak (100\% capacity)      & $\sim$500\,m & $\sim$800\,m \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis}: The Kubernetes controller consumes substantially more CPU at idle due to several baseline costs:
\begin{itemize}
    \item Continuous watch connections to the API server with periodic re-list, even when no resources change~\cite{client_go}
    \item Informer caches maintained in memory for every watched resource~\cite{controller_runtime, ibryam2020}
    \item Standard controller-runtime metrics (workqueue depth, reconcile counters) and Prometheus client overhead
\end{itemize}

\subsection{Memory Usage}

\begin{table}[h]
\centering
\caption{Approximate resident memory (MB), median over sampling intervals.}
\label{tab:memory-usage}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{State} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Idle       & $\sim$50\,MB  & $\sim$150\,MB \\
Under load & $\sim$150\,MB & $\sim$300\,MB \\
Peak       & $\sim$300\,MB & $\sim$500\,MB \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis}: The Kubernetes controller uses 3$\times$ more memory due to:
\begin{itemize}
    \item controller-runtime cache (in-memory informer stores)
    \item Kubernetes client-go overhead
    \item Larger Go runtime footprint (more goroutines)
\end{itemize}

\subsection{Startup Time}

\begin{table}[h]
\centering
\caption{Approximate startup time, in seconds. Phases are timed from process launch to the listed event.}
\label{tab:startup-time}
\small\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Phase} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Binary load                 & $\sim$0.1\,s & $\sim$0.1\,s \\
Database / API connection   & $\sim$0.5\,s & $\sim$2.0\,s \\
Migrations / CRD check      & $\sim$1.0\,s & $\sim$1.5\,s \\
Ready to serve              & $\sim$2.0\,s & $\sim$10\,s  \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis.} The controller takes substantially longer to reach readiness because it must complete the Kubernetes API-server handshake, validate the registered CRD versions, run informer cache synchronisation for both Subscription and BillingPlan resources, and acquire the leader-election lease before considering itself ready. The backend's startup is dominated by GORM auto-migration against PostgreSQL; the rest is constant.

\section{Operational Characteristics}

\subsection{Scaling}

\textbf{backend-billing}:
\begin{itemize}
    \item Manual horizontal scaling via Docker Compose replicas
    \item Requires external load balancer (nginx, HAProxy)
    \item Database connection pooling limits scale
\end{itemize}

\textbf{kube-billing}:
\begin{itemize}
    \item Automatic scaling via HorizontalPodAutoscaler
    \item Min 1, Max 5 replicas (configurable)
    \item CPU target: 80\%, Memory target: 80\%
    \item Scale down stabilization: 300 seconds
\end{itemize}

\textbf{Analysis}: Kubernetes provides superior operational automation but adds complexity. For stable workloads, manual scaling is sufficient.

\subsection{Fault Tolerance}

\textbf{backend-billing}:
\begin{itemize}
    \item Single point of failure (no built-in HA)
    \item PostgreSQL replication required for DB resilience
    \item Crash recovery: restart from last DB state
\end{itemize}

\textbf{kube-billing}:
\begin{itemize}
    \item Kubernetes self-healing (automatic pod restart)
    \item etcd replication (3+ nodes for production)
    \item Crash recovery: reconcile from etcd state
    \item PodDisruptionBudget ensures availability during updates
\end{itemize}

\textbf{Analysis}: Kubernetes provides built-in fault tolerance but requires proper cluster configuration. The traditional backend requires manual HA setup.

\subsection{Observability Coverage}

\begin{table}[h]
\centering
\small
\caption{Observability Features}
\label{tab:observability}
\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Feature} & \textbf{backend-billing} & \textbf{kube-billing} \\ \midrule
Custom business metrics & 2 & 3 \\
HTTP/Request metrics & Yes & No \\
Reconciliation metrics & No & Yes \\
Distributed tracing & Yes (HTTP + DB) & Yes (Reconcile spans) \\
Audit events & Manual logging & Kubernetes Events API \\
Status conditions & No & Yes (metav1.Condition) \\ \bottomrule
\end{tabularx}
\end{table}

\textbf{Analysis}: The Kubernetes system provides richer business-level observability (Events, Conditions) without custom code. The traditional system offers better request-level visibility (HTTP latency, DB queries).

\section{Summary of Findings}

Table~\ref{tab:summary-comparison} summarizes the comparative analysis.

Table~\ref{tab:summary-comparison} summarises the comparison along the four evaluation axes. Each cell records the system that performs better on that axis and, where the difference is quantitative, by how much; cells whose answer is qualitative state the underlying mechanism rather than assigning a score. Subjective star ratings are deliberately avoided because there is no defensible rubric for collapsing latency, throughput, and operational properties onto a single ordinal scale.

\begin{table}[h]
\centering
\small
\caption{Comparison summary. Quantitative cells cite the measured ratio; qualitative cells cite the underlying mechanism.}
\label{tab:summary-comparison}
\begin{tabularx}{\textwidth}{@{}lXX@{}}
\toprule
\textbf{Axis} & \textbf{Where the difference lies} & \textbf{Magnitude or mechanism} \\ \midrule
Static code size &
backend-billing larger (15 vs 7 non-test files; 1{,}401 vs 1{,}013 LOC) &
The traditional layered structure (handler/repo/middleware/worker) is more verbose than the reconcile-loop concentration in kube-billing. \\
Dependency footprint &
Comparable; identical direct count (9 each) &
Transitive totals are within 12\% (213 vs 190); both stacks pull in similarly heavy ecosystems. \\
Latency (subscription create) &
backend-billing lower (Table~\ref{tab:e2e-latency}) &
HTTP+\texttt{INSERT} completes in ~10\,ms p50; kube-billing reconcile-to-Active completes in ~100\,ms p50; the gap is the apiserver+etcd write path plus watch event propagation. \\
Throughput &
backend-billing higher (Table~\ref{tab:throughput}) &
Construct caveat: the two numbers measure different contracts (HTTP request servicing vs status convergence); see Section~\ref{sec:limitations}. \\
Idle CPU/memory &
backend-billing lower (Table~\ref{tab:cpu-usage}, \ref{tab:memory-usage}) &
The controller maintains watch connections and informer caches even when no work exists; the REST server only consumes resources when a request arrives. \\
Scaling &
kube-billing automated (Listing~\ref{lst:k8s-deploy}) &
HPA scales 1--5 replicas on CPU/memory triggers; the backend has no leader election in its worker, so multi-replica is unsafe without external coordination. \\
Fault tolerance &
kube-billing built-in &
Pod restart, declarative state reconstruction from etcd, and PDB-bounded voluntary disruption are properties of the platform, not of the application. \\
Cleanup semantics &
kube-billing guaranteed &
Finalizer-bounded deletion runs the pro-rated refund even on controller restart; the backend's cancel endpoint has no equivalent guarantee. \\
Audit trail &
kube-billing structured &
Kubernetes Events and metav1.Conditions are queryable by tooling; the backend's equivalent lives only in application logs. \\
Portability &
backend-billing higher &
A Docker container with PostgreSQL runs anywhere; kube-billing requires a Kubernetes API server. \\
\bottomrule
\end{tabularx}
\end{table}

\section{Discussion}

\subsection{When to Use Kubernetes Operator Approach}

The Kubernetes-based approach is advantageous when:

\begin{itemize}
    \item \textbf{Application is already on Kubernetes} --- no additional infrastructure required
    \item \textbf{High availability is critical} --- self-healing and HPA provide resilience
    \item \textbf{GitOps workflow is desired} --- declarative resources integrate with ArgoCD/Flux
    \item \textbf{Business logic is state machine-like} --- reconcile pattern fits well
    \item \textbf{Audit trail is required} --- Kubernetes Events provide built-in logging
\end{itemize}

\subsection{When to Use Traditional Backend Approach}

The traditional backend is preferable when:

\begin{itemize}
    \item \textbf{Performance is critical} --- 10--20$\times$ lower latency
    \item \textbf{Team lacks Kubernetes expertise} --- simpler development model
    \item \textbf{Application must run outside Kubernetes} --- portability requirement
    \item \textbf{Time-to-market is priority} --- faster development cycle
    \item \textbf{Complex queries are needed} --- SQL vs Kubernetes API limitations
\end{itemize}

\subsection{Threats to Validity}

The full catalogue of limitations and validity threats --- including the construct mismatch between HTTP and reconcile-driven throughput, the asymmetric state vocabularies, and the absence of failure injection --- is consolidated in Section~\ref{sec:limitations}.

\section{Summary}

This chapter presented experimental evaluation of both implementations. The traditional backend outperforms the Kubernetes Operator in raw performance (10--20$\times$ lower latency, 5$\times$ higher throughput) and resource efficiency (3--5$\times$ less CPU/memory). However, the Kubernetes approach provides superior operational characteristics: automatic scaling, self-healing, and built-in observability (Events, Conditions). The choice between approaches depends on organizational priorities: performance and simplicity favor the traditional backend, while operational automation and cloud-native integration favor the Kubernetes Operator.
