\chapter{Introduction}\label{chap:intro}

\section{Background and Context}

The rapid evolution of cloud-native computing has fundamentally transformed how modern software systems are architected, deployed, and operated. At the core of this transformation lies Kubernetes, which has emerged as the de facto standard for container orchestration since its open-source release in 2014~\cite{kubernetes, luksa2017}. Initially developed by Google based on their internal Borg system and later donated to the Cloud Native Computing Foundation (CNCF), Kubernetes provides robust capabilities for automating deployment, scaling, and management of containerized applications~\cite{cncf, hightower2017}.

However, Kubernetes offers far more than container orchestration capabilities. Its extensible architecture, built around a declarative API model and a control plane pattern, enables it to function as a universal control plane for managing arbitrary resources~\cite{kubernetes_declarative_automation}. Through Custom Resource Definitions (CRDs) and the Operator pattern, Kubernetes can be extended to manage not only infrastructure components but potentially business logic itself~\cite{crd_kubernetes, operator_pattern, dobies2020}. This extensibility has been thoroughly documented in both academic literature and industry best practices~\cite{hausenblas2019, ibryam2020}.

Modern cloud systems strive for maximum automation and flexibility. Organizations are increasingly seeking unified platforms that can manage both infrastructure and application logic through consistent interfaces~\cite{kubernetes_multivocal_review}. The declarative nature of Kubernetes, combined with its powerful extension mechanisms, positions it as a potential candidate for implementing business logic, not just orchestrating containers. This paradigm shift raises fundamental questions about the boundaries of the Infrastructure-as-Code approach and whether Kubernetes can serve as a unified control plane for both infrastructure and application concerns~\cite{morris2020iac}.

\section{Problem Statement}

Despite the widespread adoption of Kubernetes for infrastructure management, there remains a lack of systematic study regarding its applicability for implementing business logic using CRDs and the Operator pattern. Traditional application architectures typically separate business logic implementation from infrastructure concerns, with applications deployed as microservices that implement domain-specific functionality through conventional programming approaches.

Several projects have demonstrated the power of extending Kubernetes through CRDs and Operators, including Crossplane for infrastructure provisioning~\cite{crossplane}, Kubeflow for machine learning workflows, and various database operators~\cite{openshift_operators_lifecycle}. However, most existing solutions focus on infrastructure management or specialized workloads rather than general-purpose business logic~\cite{kubebuilder}.

The main problem lies in understanding whether and how Kubernetes, through CRDs and Operators, can effectively serve as a platform for implementing business logic. Key questions include: What are the performance characteristics of such an approach? How does it compare to traditional REST API-based architectures? What are the operational implications? What types of business problems are well-suited to this paradigm?

Solving this problem is important for several reasons. First, it could enable organizations to consolidate their technology stacks by using Kubernetes as a unified platform. Second, it may provide new patterns for building highly automated, self-healing business applications. Third, it could help determine clear boundaries for when this approach is beneficial versus when traditional architectures should be preferred.

\section{Research Questions}

This thesis investigates the following research questions:

\textbf{RQ1}: How can business logic be effectively modeled and implemented using Kubernetes Custom Resource Definitions and the Operator pattern?

\textbf{RQ2}: What are the performance characteristics of a business logic implementation based on CRDs and Operators compared to traditional REST API architectures?

\textbf{RQ3}: What are the operational implications and observability capabilities of using Kubernetes as a business logic platform?

\textbf{RQ4}: Under what circumstances is the CRD/Operator approach advantageous for implementing business logic, and what are its limitations?

\section{Research Objectives and Contributions}

The primary objective of this research is to explore and evaluate Kubernetes as a platform for implementing business logic through Custom Resource Definitions and the Operator pattern. Specific objectives include:

\begin{enumerate}
    \item Study existing approaches to extending the Kubernetes API through CRDs and Operators, analyzing patterns and best practices established by the cloud-native community~\cite{ibryam2020, hausenblas2019}.
    
    \item Design and develop a prototype application that implements representative business logic using CRDs and Operators, demonstrating the feasibility of this approach.
    
    \item Establish comprehensive observability for the prototype system using cloud-native tools including Prometheus and OpenTelemetry to enable detailed performance analysis~\cite{prometheus, opentelemetry, prometheus_cloud_native_monitoring}.
    
    \item Conduct experimental comparisons between the CRD/Operator-based implementation and a traditional architecture using PostgreSQL with REST API\@.
    
    \item Analyze results across multiple dimensions including reaction time to events, implementation complexity, observability capabilities, and system reliability.
\end{enumerate}

The central thesis advanced by this work is that \emph{Kubernetes Operators constitute a lifecycle-oriented business state management platform: a programming model distinct both from the request-response model of REST microservices and from the resource-provisioning focus of infrastructure operators, and one whose cost relative to a conventional REST backend is empirically quantifiable rather than anecdotal}. From this thesis, the contributions of the work are:

\begin{itemize}
    \item \textbf{Quantification of the latency--automation trade-off.} The first comparative measurement, in the open literature known to the author, of the latency, throughput, and resource-utilisation cost of running an identical business workflow as a Kubernetes Operator versus a conventional REST-and-relational-database backend, conducted on functionally comparable reference implementations~\cite{kubernetes_multivocal_review, etcd_impact}.

    \item \textbf{A reference programming model for lifecycle-oriented business logic.} A demonstration that the reconcile loop, finalizers, and Conditions form a sufficient programming surface for stateful business workflows whose natural unit of work is a resource lifecycle rather than an individual request~\cite{ibryam2020, dobies2020}.

    \item \textbf{Empirical characterisation of construct asymmetries.} An explicit account of where the two architectures cannot be aligned --- cancellation semantics, error-as-state versus error-as-Condition, audit trails --- with the asymmetries treated as findings rather than smoothed away as implementation details.

    \item \textbf{A decision framework anchored in evidence rather than intuition.} Practitioner-facing criteria for when the operator pattern is justified for business logic, anchored in the measured trade-offs rather than in folklore about cloud-native architecture~\cite{kubernetes_multivocal_review}.

    \item \textbf{Two open-source reference implementations.} Codebases that can be re-executed, modified, and extended for further research; both repositories include their benchmark harnesses and raw data.
\end{itemize}

\section{Research Methodology}\label{sec:methodology}

This research employs a mixed-methods approach combining system design, implementation, and experimental evaluation:

\textbf{Phase 1 — Literature Review and Architecture Design}: Conduct comprehensive review of existing literature on Kubernetes extensibility, CRDs, Operators, and related patterns~\cite{luksa2017, dobies2020, hausenblas2019}. Design the architecture for both the CRD/Operator-based system and the baseline traditional system.

\textbf{Phase 2 — Implementation}: Develop the prototype systems using appropriate tools and frameworks. The CRD/Operator implementation will utilize Kubebuilder for scaffolding and controller development~\cite{kubebuilder}. Both systems will implement equivalent business logic to enable fair comparison.

\textbf{Phase 3 — Observability Setup}: Configure comprehensive monitoring using Prometheus for metrics collection and OpenTelemetry for distributed tracing~\cite{prometheus, opentelemetry, prometheus_cloud_native_monitoring}. Implement GitOps practices using ArgoCD for deployment management~\cite{argocd, gitops}.

\textbf{Phase 4 — Experimental Evaluation}: Conduct controlled experiments measuring key performance indicators including reaction time to events, implementation complexity, observability coverage, and reliability~\cite{hpc_monitoring_kubernetes_prometheus}.

\textbf{Phase 5 — Analysis and Conclusions}: Analyze experimental results, synthesize findings, and formulate conclusions on the applicability of the approach.

\section{Thesis Organization}

The remainder of this thesis is organized as follows:

\textbf{Chapter~\ref{chap:lr} --- Literature Review}: Provides background on Kubernetes architecture, Custom Resource Definitions, the Operator pattern, related approaches (Infrastructure-as-Code, GitOps, service meshes), and observability tooling. Identifies the gaps that motivate this work.

\textbf{Chapter~\ref{chap:met} --- Methodology}: Describes the research approach, the chosen business domain (billing and subscription management), and the high-level architectural decisions for both the Kubernetes-based system and the traditional baseline. Outlines the evaluation framework.

\textbf{Chapter~\ref{chap:impl} --- Implementation}: Covers implementation details including CRD schemas, controller logic, observability instrumentation, deployment manifests, and the corresponding traditional REST/PostgreSQL stack.

\textbf{Chapter~\ref{chap:eval} --- Evaluation and Discussion}: Presents the experimental setup, methodology, and results of the comparative evaluation across code metrics, performance, resource utilization, and operational characteristics, together with the analysis and discussion of those results.

\textbf{Chapter~\ref{chap:conclusion} --- Conclusion}: Answers the research questions, summarizes the contributions and limitations, and suggests directions for future work.

Recurring terms (CRD, Operator, Controller, Reconciler, Spec/Status, Finalizer, Condition, RequeueAfter, Eventually consistent, GitOps) are defined together in Section~\ref{sec:terminology} so that the rest of the thesis can use them without re-introducing each on first use.

\section{Terminology}\label{sec:terminology}

The following terms recur throughout the thesis and are used with the meanings stated below.

\begin{description}
    \item[\textbf{Custom Resource Definition (CRD).}] A cluster-scoped schema, registered with the Kubernetes API server, that introduces a new resource kind alongside the built-in ones~\cite{crd_kubernetes, hausenblas2019}. The CRD names the kind, its API group and version, an OpenAPI v3 validation schema, and optionally a status subresource.

    \item[\textbf{Custom Resource (CR).}] An individual instance of a CRD --- the YAML or JSON document a user applies with \texttt{kubectl}. Where the literature uses ``CRD'' loosely to mean both schema and instance, this thesis distinguishes the two.

    \item[\textbf{Controller.}] A process that watches one or more resource kinds and reconciles their observed state against their declared state. Controllers may be built-in (the deployment controller, the service controller) or custom~\cite{controller_pattern, ibryam2020}.

    \item[\textbf{Operator.}] A custom controller paired with one or more CRDs, where the controller encodes operational knowledge specific to a domain. The terms ``operator'' and ``custom controller'' are sometimes conflated; this thesis uses ``operator'' when the controller-plus-CRDs combination is being discussed as a packaged unit, and ``controller'' when only the runtime behaviour is in scope~\cite{operator_pattern, dobies2020}.

    \item[\textbf{Reconciler.}] The function (typically named \texttt{Reconcile}) inside a controller that processes a single work-queue item. Each invocation is given the namespaced name of the resource to act on and returns a result that may schedule a future reconciliation.

    \item[\textbf{Reconciliation loop.}] The event-driven loop in which a controller observes a resource change via the Kubernetes watch API, enqueues the affected resource, and dispatches the work-queue item to the reconciler. ``Loop'' is a misnomer in that there is no polling; reconciliations are triggered by events and by explicit re-queue requests.

    \item[\textbf{Execution model.}] In this thesis, the mechanism by which a system triggers and processes business state transitions. The traditional backend's execution model is request-response with a background polling worker; the operator's execution model is event-driven reconciliation with controller-runtime-scheduled re-queueing.

    \item[\textbf{Spec and Status.}] The two halves of a Kubernetes resource. \texttt{spec} is the declared desired state, written by users or higher-level controllers; \texttt{status} is the observed state, written by the resource's controller through a dedicated subresource that bypasses optimistic locking on the spec.

    \item[\textbf{Finalizer.}] A string-valued entry in a resource's metadata that prevents deletion until removed. A controller treats the presence of a deletion timestamp on a finalizer-bearing resource as a signal to run cleanup work; on completion it removes its own finalizer entry, at which point the API server completes the deletion. Finalizers guarantee that cleanup hooks execute even across controller restarts.

    \item[\textbf{Condition.}] A standardised \texttt{metav1.Condition} structure inside \texttt{status.conditions} carrying a type, a True/False/Unknown status, a machine-readable reason, a human-readable message, and the generation observed when the condition was last set. Multiple conditions on the same resource provide a multi-dimensional state representation that a single state field cannot~\cite{ibryam2020}.

    \item[\textbf{RequeueAfter.}] A field in the result returned by \texttt{Reconcile} that asks controller-runtime to enqueue another reconciliation after a specified delay. The mechanism replaces polling: the controller does not actively check; it instructs the runtime when to wake it up.

    \item[\textbf{Admission control.}] The chain of validating and mutating webhooks that the API server consults before persisting a resource write. CRD schema validation runs as part of admission, which means invalid resources are rejected before they reach the controller.

    \item[\textbf{Eventually consistent.}] A property of the operator pattern: a controller does not commit to bringing \texttt{status} into agreement with \texttt{spec} within a bounded interval, only to retrying until success is achieved. Failures are recoverable by definition.

    \item[\textbf{GitOps.}] A deployment discipline in which the desired state of a system is captured in a Git repository and an in-cluster agent (Argo~CD, Flux) continuously reconciles the cluster toward that committed state~\cite{argocd, gitops}.
\end{description}

\section{Summary}

This chapter has introduced the research problem of using Kubernetes as a platform for business logic through Custom Resource Definitions and the Operator pattern. While Kubernetes has proven highly successful for infrastructure management, its applicability for implementing business logic remains understudied~\cite{kubernetes_multivocal_review}. This thesis aims to address this gap through systematic design, implementation, and evaluation of both CRD/Operator-based and traditional architectures. The expected outcomes include practical insights into when and how this approach can be beneficial, contributing to the broader understanding of cloud-native application development patterns.
